---
title: Overview of nimbleMacros
author: Ken Kellner
output:
  markdown::html_format
vignette: >
  %\VignetteIndexEntry{Overview of nimbleMacros}
  %\VignetteEngine{knitr::knitr}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(message = FALSE)
suppressMessages(library(nimbleMacros))
```

# Introduction

The `nimbleMacros` package has two goals:

1. Provide a set of macros for linear modeling that may be used by both end users and R package developers.
2. Demonstrate the use of macros in [NIMBLE](https://r-nimble.org/) models; and

We intend the macros in this package to be widely useful, but also just a starting point and example for NIMBLE users to write their own macros.
To begin, load the `nimbleMacros` package, which will also load the `NIMBLE` package:

```{r, eval=FALSE}
library(nimbleMacros)
```

Currently, macro functionality is disabled by default in `NIMBLE`.
Loading `nimbleMacros` will activate it automatically, but to do so yourself, you can also set the option manually:

```{r, eval = FALSE}
nimbleOptions(enableModelMacros = TRUE)
```

We'll begin by showing off the macros available in the package, and then discuss building your own macros at the end.

# `LM`: A macro for complete linear models

The `LM` macro generates complete NIMBLE model code for a variety of linear, generalized linear, and generalized linear mixed models.
It uses very similar syntax to R functions like `lm`, `glm`, `lme4::lmer`, and `lme4::glmer`.

## Example logistic regression with `mtcars`

Here's a logistic regression in R using the built-in `mtcars` dataset.
We'll model transmission type (`am`, where `am = 1` is manual transmission) as a function of standardized miles per gallon.

```{r}
rmod <- glm(am ~ scale(mpg), data = mtcars, family=binomial)
summary(rmod)
```

To fit this model with the `LM` macro, we'll start by organizing the data and constants into a list.

```{r}
const <- list(mpg = as.numeric(scale(mtcars$mpg)), am = mtcars$am)
```

Next we'll specify the prior distributions we want to use for the intercept and the slope coefficient(s).
This is done using the `setPriors` function.
We'll specify a uniform(-10, 10) distribution for the intercept and a normal distribution with mean 0 and precision 0.1 for the slope coefficient.
Note that these distribution definitions are wrapped in `quote`; you can also supply them as strings.

```{r}
pr <- setPriors(intercept=quote(dunif(-10,10)), 
                coefficient = quote(dnorm(0, 0.1)))
```

Finally, we specify the NIMBLE code, which is a single call to the `LM` macro that should look familiar to our call to `glm` above.
Two things to note:
1. We don't include a call to `scale` in the formula (`LM` does not support this). Instead we standardized the `mpg` values when we put them into our list `const`.
2. You don't need to specify dimensions of the variables with `LM`, as this is handled automatically, as `LM` assumes all data and constants are vectors of the same length. In more complex situations you can use the `LINPRED` macro; see the next section.

```{r}
code <- nimbleCode({
  LM(am ~ mpg, family = binomial, priorSpecs=pr)
})
```

Finally, we'll fit the model which yields similar results to `glm`.

```{r}

samples <- nimbleMCMC(code, constants=const, nchain=1, niter=3000, nburnin=2000,
                      samplesAsCodaMCMC = TRUE)
summary(samples)
```

## Example linear mixed model using `ChickWeights`

The `LM` macro can also generate code for random effects models.
To illustrate this we will use the built-in `ChickWeight` dataset which contains repeated measurements of chick weights over time.
First, fit the model in R using the `lme4` package:

```r
library(lme4)
rmod <- lmer(weight ~ Time + (1|Chick), data=ChickWeight)
summary(rmod)
```

```
## Linear mixed model fit by REML ['lmerMod']
## Formula: weight ~ Time + (1 | Chick)
##    Data: ChickWeight
## 
## REML criterion at convergence: 5619.4
## 
## Scaled residuals: 
##     Min      1Q  Median      3Q     Max 
## -3.0086 -0.5528 -0.0888  0.4975  3.4588 
## 
## Random effects:
##  Groups   Name        Variance Std.Dev.
##  Chick    (Intercept) 717.9    26.79   
##  Residual             799.4    28.27   
## Number of obs: 578, groups:  Chick, 50
## 
## Fixed effects:
##             Estimate Std. Error t value
## (Intercept)  27.8451     4.3877   6.346
## Time          8.7261     0.1755  49.716
## 
## Correlation of Fixed Effects:
##      (Intr)
## Time -0.422
```

To fit this model with `LM`, we'll first organize the `ChickWeight` data into a list of constants and data for NIMBLE.

```{r}
chick_const <- list(weight = ChickWeight$weight, Time = ChickWeight$Time, Chick = ChickWeight$Chick)
```

We can fit this model with `LM` using the same formula notation as `lme4` with the `(1|Chick)` part of the formula indicating random intercepts by chick.

```{r}
code <- nimbleCode({
  LM(weight ~ Time + (1|Chick))
})
```

Fit the model:

```{r}
samples <- nimbleMCMC(code, constants=chick_const, nchain=1, niter=3000, nburnin=2000,
                      samplesAsCodaMCMC = TRUE)
summary(samples)
```

# `LINPRED`

As the name suggests, the `LINPRED` macro generates a linear predictor from a provided R formula.
It is used internally by the `LM` macro and can be used separately as well.
The macro acts something like the `model.matrix` function in R, although instead of creating a model matrix, it creates the NIMBLE model code for the linear predictor.
The macro supports random effects using the same syntax as the `lme4` package.

To demonstrate the macro, we'll use the built-in `ChickWeight` dataset again.
In this dataset there are repeated weight measurements (`weight`) of chicks (`Chick`) over time (`Time`).

```{r}
head(ChickWeight)
```

## Generate a linear predictor from a formula

To create a linear predictor with time as a covariate in R, we would specify the formula as `~Time`.

```{r}
head(model.matrix(~Time, ChickWeight))
```

Based on the model matrix, the formula implies two parameters: the intercept and the coefficient/slope for time.

To generate the linear predictor NIMBLE model code, we start by setting up the constants/data list.
We'll use the `chick_const` list we used with `LM`, and add the sample size `N` so we can use it to define dimensions.

```{r}
chick_const$N <- nrow(ChickWeight)
```

In the model code, we specify the parameter name for the linear predictor (`mu`) and calculate it with the `LINPRED` macro and the same formula as above.
Here we also set an argument `priorSpecs = NULL`; we'll come back to this later.
Notice that unlike with `LM`, we explicitly define the dimensions of `mu` and `Time`.

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N], priorSpecs=NULL)
})
```

Next create the model object and view the code.

```{r}
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

NIMBLE expands the `LINPRED` macro into a `for` loop calculating a value of `mu` for each sample in the dataset,
where `mu` is a function of time.
Notice two parameters were created: `beta_Intercept` (the intercept) and `beta_Time`, the coefficient associated with time.

We can get a list of the parameters created by macros in the model using the `getMacroParameters()` function:

```{r}
mod$getMacroParameters()
```

When all the data and constants used by `LINPRED` have the same dimensions, you can specify the dimensions for the left-hand-side only:

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time, priorSpecs=NULL)
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

In more complicated situations you may have different dimensions for different data/constants.
For example, we'll convert `Time` into a one-column matrix, but keep `mu` as a vector.

```{r}
chick_const2 <- chick_const
chick_const2$Time <- matrix(chick_const2$Time, ncol=1)
str(chick_const2)
```

To handle this we just need to specify different dimensions for each in the call to `LINPRED`.

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N,1], priorSpecs=NULL)
})
mod <- nimbleModel(code, chick_const2)
mod$getCode()
```

## Factor/categorical covariates

The `LINPRED` macro can automatically handle categorical covariates (what R calls factors).
For example, suppose we included a covariate for diet type (`Diet`) in the model. 
The `Diet` covariate is already coded as an R factor:

```{r}
class(ChickWeight$Diet)
levels(ChickWeight$Diet)
chick_const$Diet <- ChickWeight$Diet
```

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N] + Diet[1:N], priorSpecs=NULL)
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

A new vector parameter `beta_Diet` is created with four elements, one per level of `Diet`.
The actual parameter `Diet` is then converted to numeric and used as the index for this vector parameter.
It isn't obvious from this code, but the first element of `beta_Diet` will be fixed at 0 (serving as the reference level).
We'll show this below.


## Add a link function

We could also add a link function to the left-hand-side of the linear predictor calculation.
Suppose we wanted to force `mu` to be positive; we could use a log link function, specifying this with the `link` argument.

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N], priorSpecs = NULL, link = log)
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

## Get default priors

If we do not manually set the `priorSpecs` argument, `LINPRED` will automatically generate a default set of priors for the two parameters it created (`beta_Intercept` and `beta_Time`).
Internally, `LINPRED` is calling the `LINPRED_PRIORS` macro (described below).

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N] + Diet[1:N])
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

We now have priors for each parameter.
Note that as mentioned previously, the first value of the `beta_Diet` vector is fixed at 0, serving as the reference level.

## Suppress the macro comments

We can also suppress the added macro comments to make things a little more concise.

```{r}
nimbleOptions(enableMacroComments = FALSE)
```

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N])
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

## Random effects

The `LINPRED` macro supports random effects using `lme4` syntax.
For example, a random intercept by `Chick`:

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N] + (1|Chick[1:N]))
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

Note that by default, all random effects are mean 0, the same approach used by `lme4`.
We can also get both random intercepts and (time) slopes by `Chick`; by default these are correlated.

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N] + (Time[1:N]|Chick[1:N]))
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

The correlated random effects model here uses an LKJ distribution prior; read more about this prior [here](https://r-nimble.org/html_manual/cha-writing-models.html#lkj-distribution-for-correlation-matrices).

As with `lme4` we can also specify uncorrelated random slopes and intercepts using `||` instead of `|`.

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N] + (Time[1:N]||Chick[1:N]))
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

## Include `LINPRED` in a complete model

Up to this point we've generated just the code needed to calculate the linear predictor.
Here's a complete linear regression model of chick weight as a function of time making use of `LINPRED` (similar to the code generated by `LM`):

```{r}
code <- nimbleCode({
  mu[1:N] <- LINPRED(~Time[1:N])

  weight[1:N] ~ FORLOOP(dnorm(mu[1:N], sd = sigma))
  sigma ~ dunif(0, 100)
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

# `LINPRED_PRIORS`

This macro generates a set of priors for parameters of a linear predictor, also based on the R formula.
The formula `~Time` implies two parameters as we described above, and `LINPRED_PRIORS` will generate a prior for each depending on the type of the parameter (e.g. intercept vs. slope).
It is not necessary to specify the dimensions in the formula when using `LINPRED_PRIORS` directly.

```{r}
code <- nimbleCode({
  LINPRED_PRIORS(~Time)
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

We can also manually set the values for different types of priors such as intercepts and coefficients (slopes).
See `?setPriors` for more details.

```{r}
pr <- setPriors(intercept = quote(dnorm(-5, 5)),
                coefficient = quote(dnorm(0, sd = 5)))

code <- nimbleCode({
  LINPRED_PRIORS(~Time + Diet, priorSpecs = pr)
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

Treating `beta_Diet` as a vector parameter makes sense in the NIMBLE model code, but it does not match the way categorical parameters are handled by `model.matrix`; i.e., dummy variables:

```{r}
colnames(model.matrix(~Time + Diet, ChickWeight))
```

Here `model.matrix` has created several separate parameters `Diet2`, `Diet3`, and `Diet4` rather than a vector.
We can replicate this structure in NIMBLE model code by setting the `LINPRED_PRIORS` option `modMatNames = TRUE`.

```{r}
code <- nimbleCode({
  LINPRED_PRIORS(~Time + Diet, priorSpecs = pr, modMatNames = TRUE)
})
mod <- nimbleModel(code, chick_const)
mod$getCode()
```

With this option set a few extra lines of NIMBLE model code are added to yield parameter names that match `model.matrix`.

# `FORLOOP`

The `FORLOOP` macro is used by `LM`, `LINPRED`, and `LINPRED_PRIORS` and is a shorthand way of creating `for` loops (possibly nested) from a single line of code.
For example, suppose you want to calculate a linear predictor `mu` from `N` values of a covariate `x`.
The NIMBLE model code would be:

```{r, eval=FALSE}
nimbleCode({
  for (i in 1:N){
    mu[i] <- alpha + beta * x[i]
  }
})
```

The `FORLOOP` macro condenses this information into a single line using index information inside parameter brackets:

```{r}
code <- nimbleCode({
  mu[1:N] <- FORLOOP(alpha + beta * x[1:N])
})
```

Here's the result after expansion, after creating some example values for the constants.

```{r}
constants <- list(x = rnorm(5), N = 5)
mod <- nimbleModel(code, constants)
mod$getCode()
```

It may seem a bit trivial to make a macro for such a simple chunk of code.
However it can be nice to use this when you have nested for loops, which are a bit more complicated.

```{r}
constants <- list(x = matrix(rnorm(10), 5, 2), N = 5, J = 2)

code <- nimbleCode({
  mu[1:N, 1:J] <- FORLOOP(alpha + beta * x[1:N, 1:J])
})

mod <- nimbleModel(code, constants)
mod$getCode()
```

Note that one place you can't use this macro directly are situations when you use the same index in multiple nested loops.
For example, suppose the parameter `mu` shown above was a square matrix.

```{r, error = TRUE}
constants <- list(x = matrix(rnorm(25), 5, 5), N = 5)

code <- nimbleCode({
  mu[1:N, 1:N] <- FORLOOP(alpha + beta * x[1:N, 1:N])
})
mod <- nimbleModel(code, constants)
```

The structure of this `for` loop with duplicated indices could be ambiguous in more complex situations, so `FORLOOP` errors.
The solution in this case would simply be to provide different variables for each dimension, even if they are the same value.

```{r}
constants <- list(x = matrix(rnorm(25), 5, 5), N = 5, K = 5)

code <- nimbleCode({
  mu[1:N, 1:K] <- FORLOOP(alpha + beta * x[1:N, 1:K])
})
mod <- nimbleModel(code, constants)
mod$getCode()
```

# Writing Your Own Macros

Suppose you wanted to create a macro called `MYDIST`, which creates a normal distribution with mean 0 and provided standard deviation.
The macro would look like this in the NIMBLE model code:

```{r, eval=FALSE}
y ~ MYDIST(sdev = 2)
```

Note that by convention, we define the macro name in all capital letters.
All the provided macros in `nimbleMacros` follow this convention.
We want the final, processed output to look like:

```{r, eval=FALSE}
y ~ dnorm(0, sd = 2)
```

## Manually creating the macro

A macro is, fundamentally, a specially-formatted R function that converts one line of code to another.
For example:

```{r}
fun <- function(input){
  LHS <- input[[2]]
  sdev <- input[[3]]$sdev
  output <- substitute(LHS ~ dnorm(0, sd = SDEV),
                       list(LHS = LHS, SDEV = sdev))
  output
}
```

Note that we have to do a bit of clunky code-processing to extract the left-hand-side of the assignment and the argument value.

```{r}
input <- quote(y ~ MYDIST(sdev = 2))
fun(input)
```

In practice, `nimble` expects this function to have two additional arguments: `modelInfo` and `.env`.
When processing a macro, `nimble` will provide a named list to the `modelInfo` argument which contains some additional model information.
Most importantly, this list includes `constants`, which can be used and/or changed by the macro.
As you can probably guess, `nimble` provides the R environment in which `nimbleModel` was called to the `.env` argument, to help with scoping issues.

In addition to these two arguments, the function also needs a bit more structure in the output.
The output should be a named list with two elements: `code`, which contains the output code (as in the original function above), and `modelInfo` which is the list of model information, which we could modify or leave as-is.

```{r}
fun <- function(input_code, modelInfo, .env){
  LHS <- input_code[[2]]
  sdev <- input_code[[3]]$sdev
  output_code <- substitute(LHS ~ dnorm(0, sd = SDEV),
                       list(LHS = LHS, SDEV = sdev))

  # Modify the constants
  modelInfo$constants$new <- 1
    
  list(code = output_code, modelInfo = modelInfo)
}
```

```{r}
input <- quote(y ~ MYDIST(sdev = 2))
fun(input, modelInfo = list(constants=NULL), .env = environment())
```

Finally, `nimble` needs a way to know that this function is intended to be a macro, and that it is called `MYDIST`.
The final step then is to create a wrapper `list` with our function inserted as an element called `process`, and assign the class `"model_macro"`.

```{r}
MYDIST <- list(process = fun)
class(MYDIST) <- "model_macro"
```

Now, we can test the macro in a NIMBLE model:

```{r}
code <- nimbleCode({
  y ~ MYDIST(sdev = 2)
})
mod <- nimbleModel(code = code, data = list(y = 1))
mod$getCode()
```

## Using `model_macro_builder` to create the macro

Manually creating the macro is fine, but gets complicated if we want to extract a lot of separate pieces of information from the code line, e.g. if our macro has many arguments.
As an alternative, we can use the `model_macro_builder` function included with `nimble` which handles some of this code processing for us.
The first argument to `model_macro_builder`, is a function structured like the one we wrote above `fun`.
The other two arguments `use3pieces` and `unpackArgs` determine how the arguments of `fun` should be structured.

First, suppose both `use3pieces` and `unpackArgs` are set to `FALSE`.
In this case, the first argument to `fun` is the entire original line of code containing the macro, i.e., `y ~ MYDIST(sdev = 2)`.
This is identical to how we wrote the function above.
In this case, all `model_macro_builder` will do is create the required macro list structure and assign the class for us.

```{r}
MYDIST2 <- model_macro_builder(fun, use3pieces=FALSE, unpackArgs = FALSE)
```

```{r}
code <- nimbleCode({
  y ~ MYDIST2(sdev = 2)
})
mod <- nimbleModel(code = code, data = list(y = 1))
mod$getCode()
```

Now we will set `use3pieces = TRUE`.
In this case, `model_macro_builder` will automatically do some initial processing on the code line for us.
Specifically, it will identify if the assignment is stochastic (`stoch`, `TRUE` if `~` and `FALSE` if `<-`), and separate and return the left-hand-side (`LHS`) and right-hand-side (`RHS`) of the assignment.
With these settings, our function's first three arguments need to be `stoch`, `LHS`, and `RHS`, in that order (plus we keep the always-required `modelInfo` and `.env`).

```{r}
fun3 <- function(stoch, LHS, RHS, modelInfo, .env){
  sdev <- RHS$sdev
  output_code <- substitute(LHS ~ dnorm(0, sd = SDEV),
                       list(LHS = LHS, SDEV = sdev))

  # Modify the constants
  modelInfo$constants$new <- 1
    
  list(code = output_code, modelInfo = modelInfo)
}
```

We do a little less code processing this time since `model_macro_builder` has already separated LHS and RHS for us.
Note that in this simple example we aren't using `stoch` because we want our output code to always be stochastic.

```{r}
MYDIST3 <- model_macro_builder(fun3, use3pieces=TRUE, unpackArgs = FALSE)
```

```{r}
code <- nimbleCode({
  y ~ MYDIST3(sdev = 2)
})
mod <- nimbleModel(code = code, data = list(y = 1))
mod$getCode()
```

Finally, consider the case where we also set `unpackArgs = TRUE`.
Now, `model_macro_builder` will dig into the right-hand-side of the macro and extract the argument values for us (in this case just `sdev`).
Thus, we need to include all these arguments explicitly as arguments in our function.

```{r}
fun4 <- function(stoch, LHS, sdev, modelInfo, .env){
  output_code <- substitute(LHS ~ dnorm(0, sd = SDEV),
                       list(LHS = LHS, SDEV = sdev))

  # Modify the constants
  modelInfo$constants$new <- 1
    
  list(code = output_code, modelInfo = modelInfo)
}
```

Even less code processing is required in this case, since `model_macro_builder` will split out the `sdev` argument for us.

```{r}
MYDIST4 <- model_macro_builder(fun4, use3pieces=TRUE, unpackArgs = TRUE)
```

```{r}
code <- nimbleCode({
  y ~ MYDIST4(sdev = 2)
})
mod <- nimbleModel(code = code, data = list(y = 1))
mod$getCode()
```

The choice of arguments will depend on the macro: sometimes the short-cuts enabled by `model_macro_builder` will be useful, other times you may need more manual control over how the input code line is processed.
See `?model_macro_builder` for additional detailed documentation and examples, including an example when `use3pieces = FALSE` and `unpackArgs = TRUE`.
You may also find the source code for the macros included in `nimbleMacros`, found [here](https://github.com/nimble-dev/nimbleMacros/tree/master/R), a useful resource for writing your own.
